### 2013-10-01 Publications/2013-ISBI14/01-abstract.tex
Manus' changes 2

\begin{abstract}
Connectivity analysis on diffusion MRI data of the whole-brain
suffers from distortions originated by the standard 
echo-planar imaging sequences used in acquisition. %
%These artifacts are present at interfaces between tissues,
%due to the inhomogeneity of the field derived from the 
% discontinuity of magnetic susceptibility.
These images show characteristic geometrical deformations
and signal destruction that are a significant pitfall
limiting the success of tractography algorithms.

>>>>

\begin{abstract}
Connectivity analysis on diffusion MRI data of the whole-brain
suffers from distortions caused by the standard 
echo-planar imaging sequences used in acquisition. %
%These artifacts are present at interfaces between tissues,
%due to the inhomogeneity of the field derived from the 
% discontinuity of magnetic susceptibility.
These images show characteristic geometrical deformations
and signal destruction that are a significant pitfall
limiting the success of tractography algorithms.

### 2013-10-01 Publications/2013-ISBI14/02-introduction.tex
Manus' changes 2

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are especially sensitive.
Magnetic susceptibility disturbs the magnetic field 
close to tissue interfaces. This inhomogeneity 
of the field translates in a highly distorted imaged 
anatomy and a significant signal destruction of 
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within the context of
functional MRI because they are also acquired with 
\gls*{epi}.

@@

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits benchmarking 
strategies. Recently, Irfanoglu et al.
\cite{irfanoglu_effects_2012} raised the question
of distortion-derived impacts in tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

>>>>

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are especially sensitive.
Magnetic susceptibility disturbs the magnetic field 
close to tissue interfaces. This inhomogeneity 
of the field translates in a highly distorted imaged 
anatomy and a significant signal destruction of 
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
profoundly described, generally within the context of
functional MRI which also uses
\gls*{epi}.

@@

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits benchmarking 
strategies. Recently, Irfanoglu et al.
\cite{irfanoglu_effects_2012} raised the question
of distortion-derived impact on tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

### 2013-10-01 Publications/2013-ISBI14/03-methods.tex
Manus' changes 2

Once a realistic \gls*{vsm} was available, we generated
the corresponding distorted \glspl*{dwi}, in two opposed
phase-gradient encoding directions. The second simulated
``acquisition'' of the same phantom was necessary 
for evaluating \gls*{reb} methods.

@@

Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework. \Gls*{fmb} correction
is essentially the same as the one used for generating the 
distortion, using the synthetic fieldmap to infer a new \gls*{vsm}.
For practical reasons (distortion on the mask, signal dropout), the
method did not achieve a perfect geometrical correction as might be
expected if the process was exactly inverse. For \Gls*{reb} correction,
we made use of the implementation found in \emph{FSL} (\texttt{topup})
that demands for the \textit{b0} of the reverse-encoding simulation.
In this second case, the \gls*{vsm} is inferred from the differences
between the two corresponding \textit{b0}.

@@

In order to enable the analysis of the influence of
distortions on the final connectivity outcome,
an intermediate step was also implemented to
produce deterministic tractography,
using the \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}.
The original phantom, one distorted version, and
the corrected instances are fed to this module.
Additionally, the original tissue probability maps
are also distorted and corrected, for computing overlap
indices.
Each tractography was restricted to its corresponding 
\gls*{wm} mask, after distortion and correction when
applied. For all the cases, we used 10 random seeds at each seeding
voxel and the seeding voxels were taken from the 
ground-truth, mimicking the usual procedure on real data
(where regions are typically mapped from the
anatomically correct T1).

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
signal similarity frame by frame computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
\glspl*{dwi}. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

>>>>

From this synthetic \gls*{vsm},
we generated
the corresponding distorted \glspl*{dwi}, in two opposed
phase-gradient encoding directions. The second simulated
``acquisition'' of the same phantom was necessary 
for evaluating \gls*{reb} methods.

@@

Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework.

\Gls*{fmb} correction
is essentially the same as the one used for generating the 
distortion, using the synthetic fieldmap to infer a new \gls*{vsm}.

For \Gls*{reb} correction, we made use of the implementation found 
in \emph{FSL} (\texttt{topup})
that demands for the \textit{b0} of the reverse-encoding simulation.
In this second case, the \gls*{vsm} is inferred from the differences
between the two corresponding \textit{b0}.

@@

To evaluate the influence of
distortions on the final connectivity outcome,
an intermediate step was also implemented to
produce deterministic tractography,
using the \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}.
The original phantom, one distorted version, and
the corrected instances are fed to this module.
Additionally, the original tissue probability maps
are also distorted and corrected, for computing overlap
indices.
Each tractography was restricted to its corresponding 
\gls*{wm} mask, after distortion and correction when
applied. For all the cases, we used 10 random seeds at each seeding
voxel and the seeding voxels were taken from the 
ground-truth, mimicking the usual procedure on real data
(where regions are typically mapped from the
anatomically correct T1).

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
similarity volume by volume computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
\glspl*{dwi}. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

### 2013-10-01 Publications/2013-ISBI14/04-results.tex
Manus' changes 2

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed significantly
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirm the 
results presented in the Table (\autoref{fig:results}, first
row). Finally, \gls*{t2b} confirmed that its correction
can not be compared, even when using the intensity compensation
based on the determinant of the deformation field.

@@

The distortion of the phantom caused 2 false positives (inexistent 
connections created by distortion) and 2 false negatives 
(connections vanished). The correction methods had an adverse impact
on connectivity. Only \gls*{reb} could recover one lost connection
but increasing the false connections count to 3.
\gls*{fmb} ranked second, with 5 false and 2 lost connections.
Finally, \gls*{t2b} presented 9 false and 3 lost connections.

For the shake of completeness, we also document the outcome
of tractography at each step of the workflow. After distortion,
we obtained 1750 tracks and a characteristic length of 
$40.64\pm8.12$mm.
\gls*{fmb} produced 1900 tracks/$40.53\pm8.07$mm., \gls*{reb}
obtained 2200 tracks/$41.15\pm7.60$mm., and 2500 
tracks/$40.60\pm7.81$mm. for \gls*{t2b}. Although track counting has
been used as a measurement of success and as a weight of connectivity 
strength, the number of tracks produced by tractography strongly rely
on the seeding performed and the smoothing caused by interpolation used
within correction methodologies.

@@

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, what could
bias results. Nonetheless, \gls*{reb} performed better that \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
Firstly, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Secondly, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.

>>>>

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed significantly
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirms the 
results presented in the Table (\autoref{fig:results}, first
row). Finally, \gls*{t2b} confirmed that its correction
can not be compared, even when using the intensity compensation
based on the determinant of the deformation field.

@@

The distortion of the phantom caused 2 false connections 
(positive connections that are not present on the ground-truth) 
and 2 false negatives (connections vanished).
The correction methods had an adverse impact
on connectivity. Only \gls*{reb} could recover one lost connection
but increasing the false connections count to 3.
\gls*{fmb} ranked second, with 5 false and 2 lost connections.
Finally, \gls*{t2b} presented 9 false and 3 lost connections.

For the sake of completeness, we also document the outcome
of tractography at each step of the workflow. After distortion,
we obtained 1750 tracks and a characteristic length of 
$40.64\pm8.12$mm.
\gls*{fmb} produced 1900 tracks/$40.53\pm8.07$mm., \gls*{reb}
obtained 2200 tracks/$41.15\pm7.60$mm., and 2500 
tracks/$40.60\pm7.81$mm. for \gls*{t2b}. Although track counting has
been used as a measurement of success and as a weight of connectivity 
strength, the number of tracks produced by tractography strongly rely
on the seeding performed and the smoothing caused by interpolation used
within correction methodologies.

@@

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, what could
bias results. For practical reasons (distortion on the mask, 
signal dropout), the \gls*{fmb} can not achieve a perfect 
geometrical correction as might be
expected if the process was exactly inverse.
Nonetheless, \gls*{reb} performed better that \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
Firstly, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Secondly, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.
